---
title: "Bayesian Spatial Models using the INLA Approximation"
author: "Corey S. Sparks, Ph.D. - University of Texas at San Antonio"
date: "May 9, 2018"
output: 
  html_document: 
    toc: yes
---

##The INLA Approach to Bayesian models
The Integrated Nested Laplace Approximation, or INLA, approach is a recently developed, computationally simpler method for fitting Bayesian models [(Rue et al., [2009](http://onlinelibrary.wiley.com/store/10.1111/j.1467-9868.2008.00700.x/asset/j.1467-9868.2008.00700.x.pdf?v=1&t=ih5b86ev&s=9078c3b0adb48d4c15bc49ae3ededc6d1cd684c5), compared to traditional Markov Chain Monte Carlo (MCMC) approaches. INLA fits models that are classified as latent Gaussian models, which are applicable in many settings (Martino & Rue, [2010](http://www.bias-project.org.uk/gmrfcourse/inla-program.pdf).  In general, INLA fits a general form of additive models such as:

$\eta = \alpha + \sum_{j=1}^{nf} f^{(j)}(u_{ij}) + \sum_{k=1}^{n\beta}\beta_k z_{ki} + \epsilon_i$

where $\eta$ is the linear predictor for a generalized linear model formula, and is composed of a linear function of some variables u, $\beta$ are the effects  of covariates, z, and $\epsilon$ is an unstructured residual (Rue et al., 2009). As this model is often parameterized as a Bayesian one, we are interested in the posterior marginal distributions of all the model parameters. Rue and Martino [(2007)](http://www.sciencedirect.com/science/article/pii/S0378375807000845) show that the posterior marginal for the random effects (x) in such models can be approximated as:

$\tilde{p}(x_i|y) = \sum_k \tilde{p}(x_i|\theta_k, y) \tilde{p}(\theta_k|y) \Delta_k$

via numerical integration (Rue & Martino, 2007; Schrodle & Held, [2011a](http://onlinelibrary.wiley.com/doi/10.1002/env.1065/full), [2011b](http://link.springer.com/article/10.1007/s00180-010-0208-2)). The posterior distribution of the hyperparameters ($\theta$) of the model can also be approximated as:

$\tilde{p}(\theta | y)) \propto \frac{p(x, \theta, y)}{\tilde{p}G(x| \theta,y)} \mid _{x} = x^*(\theta)$

, where G is a Gaussian approximation of the posterior and $x^*(\theta)$ is the mode of the conditional distribution of $p(x|\theta,y)$. Thus, instead of using MCMC to find an iterative, sampling-based estimate of the posterior, it is arrived at numerically. This method of fitting the spatial models specified above has been presented by numerous authors (Blangiardo & Cameletti, [2015](https://books.google.com/books?hl=en&lr=&id=--HuBgAAQBAJ&oi=fnd&pg=PA259&dq=Blangiardo+%26+Cameletti,+2015&ots=VSDEJ7wfM2&sig=graudrEKTevK2HR7nozmZ-Y5N0Q#v=onepage&q=Blangiardo%20%26%20Cameletti%2C%202015&f=false); Blangiardo et al., [2013](http://www.sciencedirect.com/science/article/pii/S1877584513000336); Lindgren & Rue, [2015](http://www.sciencedirect.com/science/article/pii/S2211675315000780); Martins et al., [2013](http://www.sciencedirect.com/science/article/pii/S0167947313001552); Schrodle & Held, 2011a, 2011b), with comparable results to MCMC.


Below, I show examples of using INLA to fit Bayesian regression models for areal data from Texas counties and another example of using INLA to estimate multilevel models and perform small area estimation.  


```{r libraries, results='hide'}
#library(rgdal)
library(spdep)
library(RColorBrewer)
library(lattice)
library(INLA)
library(tigris)
library(tidycensus)
library(ggplot2)
library(dplyr)

```


```{r load data}
files<-list.files("~/Google Drive/a&m_stuff/workshop_5_14_18/vita_stat/", pattern = "*.csv", full.names = T)
vital<-lapply(files, read.csv, header=T)
library(plyr)
df <- ldply(vital, data.frame)
df$cofips<-paste(substr(df$GISJOIN, 2,3), substr(df$GISJOIN, 5,7), sep="")
df<-df%>%
  filter(YEAR %in%2000:2007, STATEA==480)%>%
  mutate(births=AGWE001, deaths=AGWG001)%>%
  select(YEAR, cofips, births, deaths)
head(df)

popurl<-url("http://www2.census.gov/programs-surveys/popest/datasets/2000-2010/intercensal/county/co-est00int-tot.csv")
pops<-read.csv(popurl)
names(pops)<-tolower(names(pops))
pops<-pops%>%
  mutate(cofips = paste(sprintf(fmt = "%02d", state), sprintf(fmt = "%03d",county), sep=""))%>%
  filter(sumlev==50, state==48)

pops$struct<-1:dim(pops)[1]
pops.long<-reshape(data = pops, idvar = "cofips", varying = list(names(pops)[9:16]), direction="long", drop = names(pops)[c(2,3,4,5,6,8,17,18,19,20)], v.names = "population")
pops.long$year<-pops.long$time+1999
head(pops.long)

dat.long<-merge(pops.long, df, by.x=c("cofips", "year"), by.y=c("cofips", "YEAR"))

#v00<-load_variables(year=2000, dataset = "sf3", cache = T)
cov_dat<-get_decennial(geography = "county", state = "TX", year = 2000, sumfile = "sf3",
                       summary_var = "P001001",
                       variables = c("P007003", "P007004","P007010","P053001", "P089001", "P089002" ),
                      output = "wide")

cov_dat<-cov_dat%>%
  mutate(cofips=GEOID,pwhite=P007003/summary_value, pblack=P007004/summary_value, phisp=P007010/summary_value,medhhinc=as.numeric(scale(P053001)), ppov=P089002/P089001)


final.dat<-merge(dat.long, cov_dat, by="cofips")

rates<-aggregate(cbind(deaths, births,population)~1, final.dat,sum)
rates$dr<-rates$deaths/rates$population
rates$br<-rates$births/rates$population


final.dat$E_d<-final.dat$population*rates$dr

final.dat$E_b<-final.dat$population*rates$br
final.dat<-final.dat[order(final.dat$cofips, final.dat$year),]
final.dat$id<-1:dim(final.dat)[1]

head(final.dat)
options(scipen=999)
```

Next we make the spatial information, we get the polygons from census directly using `counties` from the `tigris` package.
```{r, results='hide'}
us_co<-counties(state = "TX", cb = T)

#In INLA, we don't need FIPS codes, we need a simple numeric index for our counties
us_co$struct<-1:dim(us_co@data[1])
nbs<-poly2nb(us_co, queen = T, row.names = us_co$struct)
mat <- nb2mat(nbs, style="B",zero.policy=TRUE)
colnames(mat) <- rownames(mat) 
mat <- as.matrix(mat[1:dim(mat)[1], 1:dim(mat)[1]])


nb2INLA("am_graph",nbs)
am_adj <-paste(getwd(),"/am_graph",sep="")
H<-inla.read.graph(filename="am_graph")
image(inla.graph2matrix(H), xlab="", ylab="", main="")

```

```{r}
library(sf)
tx<-st_as_sf(us_co)
tx%>%
  ggplot()+geom_sf()
```

```{r}
final.dat<-merge( tx,final.dat, by="struct")
```

#Model setup
- We have a count outcome (deaths and births), in counties over time, and a set of time-constant covariates.
- We have several options in the GLM framework with which to model these data, for example:
- Binomial - $$y_{ij} \sim Bin(\pi_{ij}) \text{:  } logit(\pi_{ij} ) = \beta_{0}+ x'\beta_k $$
- Poisson - $$y_{ij} \sim Pois(\lambda_{ij} E_{ij}) \text{:  } log(\lambda_{ij} ) = log(E_{ij}) + \beta_{0}+ x'\beta_k $$
- Negative Binomial - $$y_{ij} \sim \text{Neg Bin} (\mu_{ij}, \alpha, E_{ij}) \text{:  } log(\mu_{ij} ) = log(E_{ij}) + \beta_{0}+ x'\beta_k $$
 
- In addition to various zero-inflated versions of these data.

```{r}
ggplot(data = final.dat)+geom_histogram(aes(x =deaths , y=0.5*..density..))+facet_wrap(~year)+
  ggtitle(label = "Distribution of Deaths by Year", subtitle = "Texas Counties, 2000-2007")

ggplot(data = final.dat)+geom_histogram(aes(x =deaths/E_d , y=0.5*..density..))+facet_wrap(~year)+
  ggtitle(label = "Distribution of Morality Relative Risk by Year", subtitle = "Texas Counties, 2000-2007")

```

We can fit these  model using the bayesian framework with INLA. 

First, we consider the basic GLM for the mortality outcome, with out any hierarhical structure. We can write this model as a Negative Binomial model, for instance as:

$$\text{Deaths_ij} = \text{log(E_d)} + X' \beta$$

INLA will use vague Normal priors for the $\beta$'s, and we have not other parameters in the model to specify priors for. INLA does not require you to specify all priors, as all parameters have a default prior specification.

```{r}
#Model specification:
f1<-deaths~scale(pblack)+scale(phisp)+scale(ppov)+year

#Model fit
mod1<-inla(formula = f1,data = final.dat, #linear predictor - fixed effects
           family = "nbinomial", E = E_d,  #marginal distribution for the outcome, expected count
           control.compute = list(dic=T), # compute DIC or not?
           control.predictor = list(link=1)) #estimate predicted values & their marginals or not?
#model summary
summary(mod1)

```

Plot our observed vs fitted values
```{r}
plot(x= mod1$summary.fitted.values$mean, y=final.dat$deaths/final.dat$E_d , ylab="Observed", xlab="Estimated" )


```


Now we add basic nesting of rates within counties, with a random intercept term for each county. This would allow there to be heterogenity in the mortality rate for each county, over and above each county's observed characteristics. 

This model would be:

$$\text{Deaths_ij} = \text{log(E_d)} + X' \beta + u_j$$
$$u_j \sim  \text{Normal} (0 , \tau_u)$$

where $\tau_u$ here is the precision, not the variance and **precision = 1/variance.** INLA puts a log-gamma prior on the the precision by default.

```{r}
f2<-deaths~scale(pblack)+scale(phisp)+scale(ppov)+year+ #fixed effects
  f(struct, model = "iid")  #random effects
mod2<-inla(formula = f2,data = final.dat,
           family = "nbinomial", E = E_d, 
           control.compute = list(dic=T), 
           control.predictor = list(link=1))

#total model summary
summary(mod2)

```

Marginal Distributions of hyperparameters
We can plot the posterior marginal of the hyperparameter in this model, in this case $\sigma_u = 1/\tau_u$

```{r}
m2<- inla.tmarginal(
        function(x) (1/x), #invert the precision to be on variance scale
        mod2$marginals.hyperpar$`Precision for struct`)

inla.hpdmarginal(.95, marginal=m2)

plot(m2, type="l", main=c("Posterior distibution for between county variance", "- IID model -"), xlim=c(0, .1))

```


Observed vs. Fitted values
```{r}
plot(x= mod2$summary.fitted.values$mean, y=final.dat$deaths/final.dat$E_d , ylab="Observed", xlab="Estimated" )
points(x= mod1$summary.fitted.values$mean, y=final.dat$deaths/final.dat$E_d, col=2)
legend("topleft", legend = c("GLM", "GLMM(IID)"), col=c(1,2), pch=1)

```

We see a couple of things here. First, we see much closer agreement between the observed and predicted values, and we see the shrinkage of the estimates toward the mean.


```{r}
final.dat$fitted_m2<-mod2$summary.fitted.values$mean

final.dat%>%
  filter(year%in%c(2000))%>%
  mutate(qrr=cut(fitted_m2, breaks = quantile(fitted_m2, p=seq(0,1,length.out = 8))))%>%
  ggplot()+geom_sf(aes(fill=qrr))+scale_colour_brewer(palette = "RdBu" )+scale_fill_brewer(palette = "RdBu", na.value="grey")+guides(fill=guide_legend(title="Relative Risk Quartile"))+ggtitle(label="Relative Risk Quartile - IID Model, 2000")+coord_sf(crs = 102008)
  
final.dat%>%
  filter(year%in%c(2007))%>%
  mutate(qrr=cut(fitted_m2, breaks = quantile(fitted_m2, p=seq(0,1,length.out = 8))))%>%
  ggplot()+geom_sf(aes(fill=qrr))+scale_colour_brewer(palette = "RdBu" )+scale_fill_brewer(palette = "RdBu", na.value="grey")+guides(fill=guide_legend(title="Relative Risk Quartile"))+ggtitle(label="Relative Risk Quartile - IID Model, 2007")+coord_sf(crs = 102008)

library(mapview)

map1<-final.dat%>%
  filter(year%in%c(2007))%>%
  mutate(qrr=cut(fitted_m2, breaks = quantile(fitted_m2, p=seq(0,1,length.out = 8))))
clrs <- colorRampPalette(brewer.pal(8, "RdBu"))
mapView(as(map1, "Spatial"), zcol="qrr", legend=T, col.regions=clrs, map.types="OpenStreetMap")
```


Model with spatial correlation - Besag, York, and Mollie (1991) model and temporal heterogenity
$$\text{Deaths_ij} = \text{log(E_d)} + X' \beta + u_j + v_j + \gamma_t$$
Which has two random effects, one an IID random effect and the second a spatially correlated random effect, specified as a  conditionally autoregressive prior for the $v_j$'s. This is the Besag model:

$$v_j|v_{\neq j},\sim\text{Normal}(\frac{1}{n_i}\sum_{i\sim j}v_j,\frac{1}{n_i\tau})$$
and $u_j$ is an IID normal random effect, $\gamma_t$ is also given an IID Normal random effect specification, and there are now three hyperparameters, $\tau_u$ and $\tau_v$ and $\tau_{\gamma}$ and each are given log-gamma priors.

For the BYM model we must specify the spatial connectivity matrix in the random effect.

```{r}
#final.dat$year_c<-final.dat$year - 2004
f3<-deaths~scale(pblack)+scale(phisp)+scale(ppov)+
  f(struct, model = "bym", constr = T, scale.model = T, graph = mat)+
  f(year, model="iid") #temporal random effect
mod3<-inla(formula = f3,data = final.dat,
           family = "nbinomial", E = E_d, 
           control.compute = list(dic=T), 
           control.predictor = list(link=1))

#total model summary
summary(mod3)
```

```{r}
plot(y=mod3$summary.random$year_c$mean,x=unique(final.dat$year), type="l")

m3a<- inla.tmarginal(
        function(x) (1/x),
        mod3$marginals.hyperpar$`Precision for struct (iid component)`)
m3b<- inla.tmarginal(
        function(x) (1/x),
        mod3$marginals.hyperpar$`Precision for struct (spatial component)`)
m3c<- inla.tmarginal(
        function(x) (1/x),
        mod3$marginals.hyperpar$`Precision for year`)

plot(m3a, type="l", main=c("Posterior distibution for between county variance", "- IID model -"), xlim=c(0, .1), ylim=c(0,300))
lines(m3b, col="red")
lines(m3c, col="green")

inla.hpdmarginal(.95,m3a)
inla.hpdmarginal(.95,m3b)
inla.hpdmarginal(.95,m3c)

```
This indicates **very** low spatially correlated variance in these data and very low temporal heterogenity as well.

Space-time mapping of the fitted values
```{r}
final.dat$fitted_m3<-mod3$summary.fitted.values$mean

final.dat%>%
  filter(year%in%c(2000))%>%
  mutate(qrr=cut(fitted_m3, breaks = quantile(fitted_m3, p=seq(0,1,length.out = 8))))%>%
  ggplot()+geom_sf(aes(fill=qrr))+scale_colour_brewer(palette = "RdBu" )+scale_fill_brewer(palette = "RdBu", na.value="grey")+guides(fill=guide_legend(title="Relative Risk Quartile"))+ggtitle(label="Relative Risk Quartile - IID Model, 2000")+coord_sf(crs = 102008)
  
final.dat%>%
  filter(year%in%c(2007))%>%
  mutate(qrr=cut(fitted_m3, breaks = quantile(fitted_m3, p=seq(0,1,length.out = 8))))%>%
  ggplot()+geom_sf(aes(fill=qrr))+scale_colour_brewer(palette = "RdBu" )+scale_fill_brewer(palette = "RdBu", na.value="grey")+guides(fill=guide_legend(title="Relative Risk Quartile"))+ggtitle(label="Relative Risk Quartile - IID Model, 2007")+coord_sf(crs = 102008)

library(mapview)

map1<-final.dat%>%
  filter(year%in%c(2007))%>%
  mutate(qrr=cut(fitted_m3, breaks = quantile(fitted_m3, p=seq(0,1,length.out = 8))))
clrs <- colorRampPalette(brewer.pal(8, "RdBu"))
mapView(as(map1, "Spatial"), zcol="qrr", legend=T, col.regions=clrs, map.types="OpenStreetMap")
```

##Map of spatial random effects
It is common to map the random effects from the BYM model to look for spatial trends, in this case, there are not strong spatial signals:

```{r}
tx$sp_re<-mod3$summary.random$struct$mean[1:254]
tx%>%
  mutate(qse=cut(sp_re, breaks = quantile(sp_re, p=seq(0,1,length.out = 8))))%>%
  ggplot()+geom_sf(aes(fill=qse))+scale_colour_brewer(palette = "RdBu" )+scale_fill_brewer(palette = "RdBu", na.value="grey")+guides(fill=guide_legend(title="Spatial Excess Risk"))+ggtitle(label="Spatial Random Effect - BYM Model")+coord_sf(crs = 102008)

```


##Exceedence probabilities
In Bayesian spatial models that are centered on an epidemiological type of outcome, it is common to examine the data for spatial clustering. One way to do this is to examine the clustering in the relative risk from one of these GLMM models. For instance if $\theta$ is the relative risk $$\theta = exp(\beta_0 + \beta_1*x_1 + u_j)$$ from one of our Negative binomial models above. We can use the posterior marginals of the relative risk to ask $\theta \gt \theta^*$ where $\theta^*$ is a specific level of excess risk, say 50% extra or $\theta > 1.5$. If the density, or $\text{Pr}(\theta \gt \theta^*)$ is high, then there is evidence that the excess risk is not only high, but **significantly** high. 

To get the exceedence probabilites from one of our models, we can use the `inla.pmarginal()` function to ask if $\text{Pr}(\theta \gt \theta^*)$

```{r}
thetastar<-1.5#theta*
inlaprob<- unlist(lapply(mod3$marginals.fitted.values, function(X){
   1-inla.pmarginal(thetastar, X)
}))
hist(inlaprob)

```

So, we see lots of occasions where the exceedence probability is greater than .9. We can visualize these in a map.

```{r}
final.dat$exceedprob<-inlaprob

final.dat%>%
  filter(year%in%c(2000))%>%
  mutate(qrr=cut(exceedprob, breaks = c(0, .5, .9, .95, .99, 1), include.lowest = T))%>%
  ggplot()+geom_sf(aes(fill=qrr))+scale_colour_brewer(palette = "Blues" )+scale_fill_brewer(palette = "Blues", na.value="grey")+guides(fill=guide_legend(title=""))+ggtitle(label=expression(paste("Exceedence Probability Relative Risk ","Pr( ",theta," >1.5"," )  - 2000") ))+coord_sf(crs = 102008)
  
final.dat%>%
  filter(year%in%c(2007))%>%
  mutate(qrr=cut(exceedprob, breaks = c(0, .5, .9, .95, .99, 1), include.lowest = T))%>%
  ggplot()+geom_sf(aes(fill=qrr))+scale_colour_brewer(palette = "Blues" )+scale_fill_brewer(palette = "Blues", na.value="grey")+guides(fill=guide_legend(title="Relative Risk Quartile"))+ggtitle(label=expression(paste("Exceedence Probability Relative Risk ","Pr( ",theta," >1.5"," )  - 2007") ))+coord_sf(crs = 102008)

library(mapview)

map1<-final.dat%>%
  filter(year%in%c(2007))%>%
  mutate(qrr=cut(exceedprob, breaks = c(0, .5, .9, .95, .99, 1), include.lowest = T))
  
clrs <- colorRampPalette(brewer.pal(6, "Blues"))
mapView(as(map1, "Spatial"), zcol="qrr", legend=T, col.regions=clrs, map.types="OpenStreetMap")

```

Which shows several areas of the state where risk the mortality rate is higher than the state rate.


#Multi - Level Models
```{r load data&recode, message=FALSE, warning=FALSE}
#load brfss
library(car)
library(knitr)
brfssurl<-"https://github.com/coreysparks/data/blob/master/brfss_14.Rdata?raw=true"
load(url(brfssurl))
set.seed(12345)
#samps<-sample(1:nrow(brfss_14), size = 90000, replace=F)
#brfss_14<-brfss_14[samps,]
#The names in the data are very ugly, so I make them less ugly
nams<-names(brfss_14)
#we see some names are lower case, some are upper and some have a little _ in the first position. This is a nightmare.
newnames<-gsub(pattern = "x_",replacement =  "",x =  nams)
names(brfss_14)<-tolower(newnames)
#BMI
brfss_14$bmi<-ifelse(is.na(brfss_14$bmi5)==T, NA, brfss_14$bmi5/100)
brfss_14$obese<-ifelse(brfss_14$bmi>30,1,0)
#Poor or fair self rated health
#brfss_14$badhealth<-ifelse(brfss_14$genhlth %in% c(4,5),1,0)
brfss_14$badhealth<-recode(brfss_14$genhlth, recodes="4:5=1; 1:3=0; else=NA")
#race/ethnicity
brfss_14$black<-recode(brfss_14$racegr3, recodes="2=1; 9=NA; else=0")
brfss_14$white<-recode(brfss_14$racegr3, recodes="1=1; 9=NA; else=0")
brfss_14$other<-recode(brfss_14$racegr3, recodes="3:4=1; 9=NA; else=0")
brfss_14$hispanic<-recode(brfss_14$racegr3, recodes="5=1; 9=NA; else=0")
brfss_14$race_eth<-recode(brfss_14$racegr3, recodes="1='nhwhite'; 2='nh black'; 3='nh other';
                         4='nh multirace'; 5='hispanic'; else=NA", as.factor = T)
brfss_14$race_eth<-relevel(brfss_14$race_eth, ref = "nhwhite")
#insurance
brfss_14$ins<-ifelse(brfss_14$hlthpln1==1,1,0)

#income grouping
brfss_14$inc<-ifelse(brfss_14$incomg==9, NA, brfss_14$incomg)

#education level
brfss_14$educ<-recode(brfss_14$educa, recodes="1:2='0Prim'; 3='1somehs'; 4='2hsgrad';
                     5='3somecol'; 6='4colgrad';9=NA", as.factor=T)
#brfss_14$educ<-relevel(brfss_14$educ, ref='0Prim')

#employment
brfss_14$employ<-recode(brfss_14$employ, recodes="1:2='Employed'; 2:6='nilf';
                       7='retired'; 8='unable'; else=NA", as.factor=T)
brfss_14$employ<-relevel(brfss_14$employ, ref='Employed')

#marital status
brfss_14$marst<-recode(brfss_14$marital, recodes="1='married'; 2='divorced'; 3='widowed';
                      4='separated'; 5='nm';6='cohab'; else=NA", as.factor=T)
brfss_14$marst<-relevel(brfss_14$marst, ref='married')

#Age cut into intervals
brfss_14$agec<-cut(brfss_14$age80, breaks=c(0,24,39,59,79,99), include.lowest = T)

```

I want to see how many people we have in each MSA in the data:
```{r}

#Now we will begin fitting the multilevel regression model with the msa
#that the person lives in being the higher level
head(data.frame(name=table(brfss_14$mmsaname),id=unique(brfss_14$mmsa)))
#people within each msa

#How many total MSAs are in the data?
length(table(brfss_14$mmsa))
#counties
```


###Higher level predictors
We will often be interested in factors at both the individual *AND* contextual levels. To illustrate this, I will use data from the American Community Survey measured at the MSA level. Specifically, I use the DP3 table, which provides economic characteristics of places, from the 2010 5 year ACS [Link](http://www.census.gov/acs/www/data_documentation/special_data_release/).
```{r, echo=FALSE}
mykey<-"997fb9115102b709d5028501b4b030e84af62525"
```


```{r load_acs,message=FALSE, warning=FALSE}
library(acs)
#Get 2010 ACS median household incomes for tracts in Texas
msaacs<-geo.make(msa="*")

acsecon<-acs.fetch(key=mykey, endyear=2010, span=5, geography=msaacs, variable = c("B19083_001","B17001_001","B17001_002", "B03002_001","B03002_004", "B03002_012" ))

colnames(acsecon@estimate)

msaecon<-data.frame(gini=acsecon@estimate[, "B19083_001"], 
ppoverty=acsecon@estimate[, "B17001_002"]/acsecon@estimate[, "B17001_001"],
pblack=acsecon@estimate[,"B03002_004"]/acsecon@estimate[, "B03002_001"],
phisp=acsecon@estimate[,"B03002_012"]/acsecon@estimate[, "B03002_001"],
giniz=scale(acsecon@estimate[, "B19083_001"]), 
ppovertyz=scale(acsecon@estimate[, "B17001_002"]/acsecon@estimate[, "B17001_001"]))
msaecon$zpblack<-scale(msaecon$pblack)
msaecon$zphisp<-scale(msaecon$phisp)
msaecon$ids<-paste(acsecon@geography$metropolitanstatisticalareamicropolitanstatisticalarea)
```
Let's see the geographic variation in these economic indicators:
```{r, results='hide'}
library(tigris)
msa<-core_based_statistical_areas(cb=T)
msa_ec<-geo_join(msa, msaecon, "CBSAFP", "ids", how="inner")

tx_ec<-msa_ec[grep(msa_ec$NAME, pattern = "TX"), ]
library(RColorBrewer)
library(sp)
spplot(tx_ec, "gini", at=quantile(tx_ec$gini), col.regions=brewer.pal(n=6, "Reds"), col="transparent", main="Gini Coefficient")
spplot(tx_ec, "phisp", at=quantile(tx_ec$phisp), col.regions=brewer.pal(n=6, "Reds"), col="transparent", main="Percent Hispanic")
```

#Create spatial information for higher level units

```{r}
#See what counties are in the brfss data
tx_ec$struct<-1:dim(tx_ec)[1]
city.est.dat<-tx_ec@data[,c( "giniz","ppovertyz", "zpblack", "zphisp", "struct")]
city.est.dat$obese<-NA

head(city.est.dat)

brfss_14$cbsa<-as.character(brfss_14$mmsa)
indat<-merge(brfss_14, tx_ec, by.x="cbsa", by.y="CBSAFP", all.x=T)

brf.est<-indat[, c("giniz","ppovertyz", "zpblack", "zphisp", "struct", "obese")]
brf.est<-brf.est[order(brf.est$struct),]
head(brf.est)

##Here is where I add the cities that need to be estimated to the rest of the data
m.est<-rbind(city.est.dat, brf.est)

struct.in<-unique(brf.est$struct)
m.est$comp<-ifelse(m.est$struct%in%struct.in ,1,0)
m.est$rm<-ifelse(m.est$comp==1&is.na(m.est$obese)==T,1,0)

m.est<-m.est[-which(m.est$rm==1),]
m.est<-m.est[is.na(m.est$struct)==F,]
m.est<-m.est[order(m.est$struct),]

# 
# fake_dat<-expand.grid(race_eth=levels(brfss_14$race_eth), agec=levels(brfss_14$agec), CBSAFP=levels(as.factor(tx_ec$CBSAFP) ))
# fake_dat<-merge(fake_dat, tx_ec, by="CBSAFP")

library(spdep)
nbs<-knearneigh(coordinates(tx_ec), longlat = T, k = 4)
nbs<-knn2nb(nbs, row.names = tx_ec$struct, sym = T)
mat <- nb2mat(nbs, style="B",zero.policy=TRUE)
colnames(mat) <- rownames(mat) 
mat <- as.matrix(mat[1:dim(mat)[1], 1:dim(mat)[1]])


```


```{r}
fit_est<- inla(obese~ giniz+zpblack+zphisp+
                 f(struct, model="bym", graph=mat,constr=TRUE,  scale.model=TRUE),
               family = "binomial",Ntrials = 1,
               data=m.est,  num.threads = 2,
               control.predictor = list(link=1))
#               control.inla=list(strategy='gaussian'))
#
summary(fit_est)

m.est$est.obese<-fit_est$summary.fitted.values$mean
m.est.est<-tapply(m.est$est.obese, m.est$struct, mean, na.rm=T)
m.est.est<-data.frame(struct=names(unlist(m.est.est)), obeseest=unlist(m.est.est))
#m.est<-m.est[is.na(m.est$bmi)==T,]
msa.est<-merge(tx_ec, m.est.est, by.y="struct", by.x="struct", all.x=T, sort=F)
head(msa.est@data)

library(mapview)
clrs <- colorRampPalette(brewer.pal(8, "Blues"))

mapview(msa.est,"obeseest", col.regions=clrs, map.types="OpenStreetMap")


```

###Multi-level model in INLA
INLA is very good at doing multi-level models, and much faster than other modeling strategies. Plus we can build in the correlated random effects, such as the Besag model at the higher level of analysis. Below, we fit three models. 
A basic random intercept model:
This corresponds to a  multilevel logistic model with a higher level variables as predictors and can be written:
$$logit(y_{ij}) = \beta_{0j} + \sum {\beta x_i}$$

$$\beta_{0j} = \beta_0 + u_j$$ 

with
$$u_j \sim N(0, \tau_u)$$


```{r}
library(INLA)
indat<-indat[is.na(indat$struct)==F,]
fit_in1<- inla(obese~ race_eth+agec+f(struct, model="iid"),
               family = "binomial", Ntrials =1,
               data=indat, num.threads = 2)
summary(fit_in1)
1/fit_in1$summary.hyperpar

m<- inla.tmarginal(
        function(x) (1/x),
        fit_in1$marginals.hyperpar$`Precision for struct`)

inla.hpdmarginal(.95, marginal=m)

plot(m, type="l", main=c("Posterior distibution for between MSA variance", "- Random Intercept model -"))

``` 

That is our individual level, random intercept model. Now I will fit a model that includes MSA level demographic characteristics, the Gini index, %black and %hispanic. This corresponds to a multi-level model with higher level predictors:

$$y_{ij} = \beta_{0j} + \sum {\beta x_i} + \sum {\gamma z_j}$$

$$\beta_{0j} = \beta_0 + \sum {\gamma z_j}+ u_j$$ 

with
$$u_j \sim N(0, \tau_u)$$

```{r}
fit_in2<- inla(obese~ race_eth+agec+giniz+zpblack+zphisp+f(struct, model="iid"),
               data=indat,
               family="binomial", Ntrials = 1)
summary(fit_in2)
1/fit_in2$summary.hyperpar

m2<- inla.tmarginal(
        function(x) (1/x),
        fit_in2$marginals.hyperpar$`Precision for struct`)

inla.hpdmarginal(.95, marginal=m2)

plot(m2, type="l", main=c("Posterior distibution for between MSA variance", "- Multi-level model -"))

```

Finally, we model the county level means using the Besag-York and Mollie model that we had used on the areal data last week. In this example, we are modeling the correlation between MSAs at the higher level, instead of like last week, where we were modeling the correlation among neighboring counties.

This is a multi level model with correlation between neighboring MSAs, assuming a [Besag, York and Mollie, 1991](https://link.springer.com/article/10.1007/BF00116466)  convolution prior for the random effects
$$y_{ij} = \beta_{0} +\sum {\beta x_i} + \sum {\gamma z_j}  + u_j + v_j$$

$$\beta_{0j} = \beta_0 + \sum {\gamma z_j}+ u_j+v_j$$ 

with

$$u_j \sim N(0, \tau_u)$$

$$v_j|v_{\neq j},\sim\text{Normal}(\frac{1}{n_i}\sum_{i\sim j}v_j,\frac{1}{n_i\tau})$$


```{r}

fit_in3<- inla(obese~ race_eth+agec+giniz+zpblack+zphisp+
                 f(struct, model="bym", graph=mat),
               family = "binomial", Ntrials = 1, 
               data=indat,
               control.predictor = list(link=1))

summary(fit_in3)

1/fit_in3$summary.hyperpar

m3_sp<- inla.tmarginal(
        function(x) (1/x),
        fit_in3$marginals.hyperpar$`Precision for struct (spatial component)`)
m3_iid<- inla.tmarginal(
        function(x) (1/x),
        fit_in3$marginals.hyperpar$`Precision for struct (iid component)`)

inla.hpdmarginal(.95, marginal=m3_sp)
inla.hpdmarginal(.95, marginal=m3_iid)

plot(m3_sp, type="l", main=c("Posterior distibution for between Spatial MSA variance", "- Multi-level model -"), xlim=c(0, .015))
lines(m3_iid, col=2,lty=2)
legend("topright", legend=c("Spatial Variance", "IID Variance"), col=c(1,2), lty=c(1,2))

```





##References
Besag, J., York, J., & Mollie, a. (1991). Bayesian image-restoration, with 2 applications in spatial statistics. Annals of the Institute of Statistical Mathematics, 43(1), 1-20. https://doi.org/10.1007/BF00116466