---
title: "Demography Informal Methods Seminar - Introduction to Time Series Analysis"
author: "Corey Sparks, Ph.D. - UTSA Department of Demography"
date: "7/14/2020"
output:
  html_document: 
    toc: yes
    includes:
      in_header: logo.html
  pdf_document: 
    latex_engine: xelatex
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Time Series Analysis

I would suggest you read chapters 1 through 3 of [Time Series Analysis and Its Applications  4th ed](https://utsacloud-my.sharepoint.com/:b:/g/personal/corey_sparks_utsa_edu/Ear7EgBI0FVJmN-GhNqe0mIBNI5bNdB5QVZzeY2M1dnZUg?e=cA8e68) to get a introductory treatment of this topic.

In time series analysis, we are typically interested in describing the pattern of variation of a variable over time. This can be extended to include multiple variables, but the largest number of applications involve a single variable.

Time series data are very different from other data you may have encountered. These are *NOT* the same as longitudinal data, nor are they the same as event history data. Sometimes these terms are thrown around as if they are the same, but they are not.

Much of time series analysis is decriptive, in that we are wanting to describe a few basic qualities of a time series. 

In these examples, we will use the Demographic and Health Survey [Model Data](https://dhsprogram.com/data/Model-Datasets.cfm). These are based on the DHS survey, but are publicly available and are used to practice using the DHS data sets, but don't represent a real country.

In this example, we will use the outcome of contraceptive choice (modern vs other/none) as our outcome.

```{r}
library(haven)
dat<-url("https://github.com/coreysparks/data/blob/master/ZZIR62FL.DTA?raw=true")
model.dat<-read_dta(dat)

```

