---
title: "Demography Informal Methods Seminar - Population Projection and the Lee-Carter Model"
author: "Corey Sparks, Ph.D. - UTSA Department of Demography"
date: "7/21/2020"
output: html_document
---

## Population Projection
 - One of the most fundamental forms of demographic analysis
 - Uses age-structured rates of fertility and mortality to project the population structure forward into time
 - Shows patterns of population growth and age composition in future populations
 - Further analysis can show population growth rates and sensitivity of the growth rate to the vital rates
 
### Methods
 * Cohort component method 
  + $$P_{t+n} = P_t + B_t - D_t + M_t$$
  + [Example](https://www.measureevaluation.org/resources/training/online-courses-and-resources/non-certificate-courses-and-mini-tutorials/population-analysis-for-planners/lesson-8/lesson-8-the-cohort-component-population-projection-method)
 
 * Hamilton-Perry method
  + Very low data requirements
  + Uses cohort change ratios from two previous censuses to project the population forward
  + $$P_{t+n} = CCR_{t, t-n} * P_t$$
  Where the $CCR$ is the ratio of the population age structure at the two previous censuses.
  + Good description is [here](https://demographics.coopercenter.org/sites/demographics/files/2019-08/VAPopProj%20Methodology_2019_Updated.pdf) and [here](https://demographics.texas.gov/Resources/Presentations/ADC/2014/ADC2014_2C_Swanson.pdf)
  + Original article [here](https://utsacloud-my.sharepoint.com/:b:/g/personal/corey_sparks_utsa_edu/EVRxaOrCh21FikL16bWklzkBJly961agP92K5j2JKvk-qA?e=a1eBaT)
  
 * Leslie Matrix model [Leslie, 1945](https://www.jstor.org/stable/pdf/2332297.pdf?casa_token=jl7ggpaUHlgAAAAA:nBBAiLiNzOoobVn2vsrfnir6V0-0qYvwlDnmlCl8ofL-gXzNqZu6Q2hA5yNRBG3NluuZ3kQzZbRq1mELaDJO4Eyr9Wbmod9h_K6_rlmhFX4ZFNpJTG4)
  + Birth Pulse model
  + People reproduce at specific ages after surviving to that age
  + Very thorough treatment in [Keyfitz and Caswell, 2005, Chapters 3, 7 and 9](https://utsacloud-my.sharepoint.com/:b:/g/personal/corey_sparks_utsa_edu/EYFWQYGEoXBCukP3O75Xj1kBBmnky2P9hLFbzCbWFZ-EOQ?e=DapuF5)

$$\begin{pmatrix}
n_1\\ 
n_2\\ 
n_3
\end{pmatrix} (t+1)=  
\begin{pmatrix}
F_1 & F_2 & F_3\\ 
P_1 & 0 & 0\\ 
0 & P_2 & 0
\end{pmatrix}
\begin{pmatrix}
n_1\\ 
n_2\\ 
n_3
\end{pmatrix}(t)$$

Or as:
$$n(t+1) = \mathbf{A} n(t)$$

$n_k$ are the population sizes, $F_k$ are the reproductive values at each age, and $P_k$ are the survivorship ratios at each age

  * Very flexible - accomodates both age and stage structure, more general population model than cohort component
  * [Notes by James Holland Jones ](https://web.stanford.edu/~jhj1/teachingdocs/Jones-Leslie1-050208.pdf)
  
  * Bayesian projection methods
    + [Bayespop](https://bayespop.csss.washington.edu/)
    + Uses methods from Bayesian statistics to project TFR, $e_0$ and population structure 
    + Incorporates uncertainty in all levels of analysis
    + Leads to projections with errors incorporated
    + Used by the [UN Population Division ](https://www.un.org/development/desa/pd/#:~:text=The%20Population%20Division%20of%20the,area%20of%20population%20and%20development.)

## Example - Leslie Matrix Model
Below, I will illustrate how to use data from the [Human Mortality Database](https://www.mortality.org/) and the [Human Fertility Database](https://www.humanfertility.org/cgi-bin/main.php) for the US to create a Leslie Matrix model.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(HMDHFDplus)
library(tidycensus)
```
### Data from HMD/HFD
The human mortality and human fertility databases are excellent sources for national historic series of mortality and fertility rates. You need to register with them to get access to the data.

```{r, echo=FALSE}
mypassword="1429845433"
myusername="corey.sparks@utsa.edu"
```

```{r}
UShmd<- readHMDweb(CNTRY = "USA", item ="fltper_5x5", password =mypassword,username = myusername)
#49 for item, 51 for US

us_mort<-UShmd%>%
  filter(Year == 2015)

#average qx and lx for ages 0 to 1 and 1 to 4
us_mort$qx[us_mort$Age==0]<-us_mort$qx[us_mort$Age==0]+us_mort$qx[us_mort$Age==1]
us_mort$Lx[us_mort$Age==0]<-us_mort$Lx[us_mort$Age==0]+us_mort$Lx[us_mort$Age==1]

us_mort<-us_mort[-2,]
  
head(us_mort)
```

```{r, echo=FALSE}
myotherpassword="47515"
```

Fertility data are by single year of age, so I aggregate to 5 year intervals
```{r}
UShfd<-readHFDweb(CNTRY="USA",username = myusername,password =myotherpassword, item = "asfrRR")
#30
us_fert<-UShfd%>%
  filter(Year==2015)%>%
  mutate(age_grp = cut(Age, breaks = seq(10, 55, 5), include.lowest = T))%>%
  group_by(age_grp)%>%
  summarise(asfr5 = sum(ASFR))%>%
  filter(is.na(age_grp)==F)%>%
  ungroup()
us_fert$age5<-seq(10,50,5)

head(us_fert)

```

Combine these together
```{r}
us_dem<-merge(us_mort, us_fert, by.x="Age", by.y="age5", all.x=T)
us_dem$asfr5<-ifelse(is.na(us_dem$asfr5)==T, 0, us_dem$asfr5)

head(us_dem)

```

```{r}
ggplot(us_dem)+geom_line( aes(x=Age, y=asfr5), color="red")+geom_line( aes(x=Age, y=qx), color="blue")+xlim(0, 110)+ylab("Rate")+xlab("Age")+theme_minimal()

```

## Population data
I get the 2015 population age distribution from the census estimates using `tidycensus`
You can see the various parameters [here](https://www.census.gov/data/developers/data-sets/popest-popproj/popest/popest-vars.html)
```{r}
us_popn<-get_estimates(geography="us", product = "characteristics", breakdown = c("AGEGROUP", "SEX"), year = 2015)
us_popn<-us_popn%>%
  filter(AGEGROUP>0&AGEGROUP<19, SEX==2)%>%
  #mutate(case_when(.$AGEGROUP==1))%>%
  arrange(AGEGROUP)

head(us_popn)
```


### Leslie Matrix Construction
* Need matrix $A$, that is # of ages by # of ages in size, in this case, it will be `r dim(us_popn)[1]` x `r dim(us_popn)[1]`
* The first row are the reproductive values, $F_k$ **THESE ARE NOT AGE SPECIFIC FERTILITY RATES**, they also have to incorporate the probability of surviving to the age of reproduction. 

$$F_k = l(.5) \left ( \frac{m_i+P_i m_{i+1}}{2} \right)$$

* The *sub-diagonal is the survival ratios, these are calculated as $L_{x+1}/L_{x})$

* Another note is that this is a one-sex population model, in this case, for females only. 

* I use code from [Eddie Hunsinger's Leslie Matrix Code](https://raw.githubusercontent.com/AppliedDemogToolbox/Hunsinger_BasicProjection/master/BasicProjectionCode.R), so the good ideas here belong to him. 

```{r}

A<-matrix(0, nrow=dim(us_popn)[1], ncol=dim(us_popn)[1])
A
size<-dim(A)[1]
sxf1<-array(0,c(size-1))
Lxf<-us_dem$Lx/100000
for (i in 1:size-1){sxf1[i]<-(Lxf[i+1]/Lxf[i])}

#make matrix with survivals on off diagonal
sf1<-rbind(0,cbind(diag(sxf1),0))

##SPECIAL CALCULATION FOR OPEN-ENDED AGE GROUP OF LESLIE MATRICES
sf1[size,size]<-sf1[size,size-1]<-Lxf[size]/(Lxf[size]+Lxf[size-1])

# proportion of female births
ffab<-0.4882846

##MAKE THE LESLIE MATRICES FOR FEMALES
#convert fertilities to proportions
fert<-us_dem$asfr5#/sum(us_dem$asfr5)

##Make first row of matrix - reproductive values
for(j in 1:size-1)
{sf1[1,j]<-(Lxf[1]/10)*(fert[j]+fert[j+1]*(sxf1[j]))*ffab}
sf1

#assemble matrix
sf1[size,size]<-0
A<-sf1
plot(x=seq(0, 80, 5), y=diag(sf1[-1,-ncol(sf1)]), type="l", ylim=c(0, 1), ylab="Rate", xlab="Age")
lines(x=seq(0, 80, 5), y = sf1[1, -size], col=2)
```

## Project the population
Since these data represent 5 year age intervals, the projections will move the population forward in increments of 5 years. 
Below, I project the population forward by 10 periods, or 50 years. Given that we start at 2015 using data from the HMD/HFD, this will take us to 2065.

The weakness of this model is that it does not incorporate migration, so this is an incomplete model, but reflects the extrapolation of the population using current and unchanging rates of fertility and mortality. The model can be expanded to incorporate segmented populations, however, with exchanges between areas. 

```{r}
##MAKE ARRAYS TO HOLD THE DATA
nproj<-10
newpop<-matrix(0, nrow=size, ncol=nproj)
newpop[,1]<-us_popn$value #first column is current population size

#loop over the new periods
for(i in 2:nproj){
newpop[,i]<-(A%*%newpop[,i-1])}

head(newpop)

#Plot the new total population sizes
options(scipen=999)
plot(y = apply(newpop, 2, sum), x= seq(2015, 2060, 5) , main="Total Female Population Size to 2060")
```


### Further analysis of the matrix...
See [Jamie's Notes](https://web.stanford.edu/~jhj1/teachingdocs/Jones-Leslie1-050208.pdf) and [Ch 13 of Keyfitz and Caswell](https://utsacloud-my.sharepoint.com/:b:/g/personal/corey_sparks_utsa_edu/EYFWQYGEoXBCukP3O75Xj1kBBmnky2P9hLFbzCbWFZ-EOQ?e=DapuF5) for more details on this.

We can do a eigenvalue decomposition of the matrix $A$, and recover the population growth rate from the log of the first eigenvalue. The *stable age structure* can also be recovered by the standardized first eigenvector of the matrix. 

```{r}
e<-eigen(A)
e$values[1]
log(e$values[1])
arv<-abs(Re(e$vectors[,1]))
stableage<-arv/sum(arv)
plot(stableage, ylim=c(0, .1))
```


Even more analysis....

```{r}
library(demogR)
A2<-leslie.matrix(lx=us_dem$lx/100000, mx=us_dem$asfr5, one.sex = T, SRB = 1.048, infant.class = F)
ea<-eigen.analysis(A2)
ea
plot(ea$stable.age)
plot(ea$sensitivities)
plot(ea$elasticities)
plot(ea$repro.value)
```

While this may seem strange, there are [other analyses](https://www.census.gov/content/dam/Census/library/publications/2020/demo/p25-1146.pdf) that show a decline in the US population in the absence of migration. Also see [this report](https://www.washingtonpost.com/local/us-population-will-decline-faster-without-steady-immigration-census-report-says/2020/02/13/1ccff6d6-4ea7-11ea-b721-9f4cdc90bc1c_story.html) in the Washington Post.

## Lee-Carter Method for Mortality Forecasting
This method originates with [Lee & Carter, 1992](https://www.tandfonline.com/doi/pdf/10.1080/01621459.1992.10475265?casa_token=eyudM2tuE8EAAAAA:FMWxjbOeaJ2NzvSQI5QkYO7xD4WA1UCIFhc47DypkjFDOsYoa2_8WMpSQNffz-v-WgwQoHYv8d-y)

They describe a method of forecasting age-specific mortality rates. Their method takes a series of age-specific mortality rates and writes it as a decomposition into an age specific average mortality, a vector of age specific changes in mortality and a period level vector of mortality trends.

$$ln(m_{x,t} = \alpha_x + \beta_x k_t +\epsilon_{x,t})$$

The solution to this equation comes from an eigenvalue decomposition of the $m_{x,t}$ matrix. Several additions and extensions of the method have been published over the years, and is an active area of research. 

```{r}
library(demography)
library(StMoMo)
library(fds)
```

## data
Again, get data from the HMD, here for the US
**Don't use my password!!**

```{r}
usdat<-hmd.mx(country = "USA", username = myusername, password = mypassword, label="USA")

usdat<-extract.years(usdat, years=1940:2017)

#male data
usdatm<-demogdata(data=usdat$rate$male,pop=usdat$pop$male,ages = usdat$age,years=usdat$year,label = usdat$label,name="male", type="mortality")

#Female data
usdatf<-demogdata(data=usdat$rate$female,pop=usdat$pop$female,ages = usdat$age,years=usdat$year,label = usdat$label,name="female", type="mortality")

summary(usdatf)
```


### Fit Lee - Carter model
I use the highest age as 90
```{r}

#males
lca1<-lca(usdatm, max.age=90)
summary(lca1)
plot(lca1)

plot(x=lca1$age, y=log(lca1$male[,1]), type="l", lty=1,
     main="Observed vs fitted Lee Carter Model - 1940 and 2017 Male Mortality",
     ylim=c(-10, 0))
lines(lca1$age, y=lca1$fitted$y[,1],
      col=1, lty=2)

lines(x=lca1$age, y=log(lca1$male[,78]), col=3)
lines(lca1$age, y=lca1$fitted$y[,78],
      col=3, lty=2)
legend("top", legend=c("Obs 1940", "Pred 1940", "Obs 2017", "Pred 2017"), col=c(1,1,3,3), lty=c(1,2,1,2))

```

```{r}
#females
lca2<-lca(usdatf, max.age=90)
summary(lca2)
plot(lca2)
plot(x=lca2$age, y=log(lca2$female[,1]), type="l", lty=1,
     main="Observed vs fitted Lee Carter Model - 1940 and 2017 Female Mortality",
     ylim=c(-10, 0))
lines(lca2$age, y=lca2$fitted$y[,1],
      col=1, lty=2)
lines(x=lca2$age, y=log(lca2$female[,78]), col=3)
lines(lca2$age, y=lca2$fitted$y[,78],
      col=3, lty=2)
legend("top", legend=c("Obs 1940", "Pred 1940", "Obs 2017", "Pred 2017"), col=c(1,1,3,3), lty=c(1,2,1,2))

#lca1<-fdm(usdatm, max.age=90, order = 3)
#lca2<-fdm(usdatf, max.age=90)

```


```{r, fig.height=8, fig.width=9}
#par(mfrow=c(1,2))
plot(usdatm,years=1940:2017,ylim=c(-11,1), ages = 60:90, main="Males - Ages 60 - 90")
out1<-forecast(lca1,h=20)

plot(usdatf,years=1940:2017,ylim=c(-11,1), ages = 60:90, main="Females - Ages 60 - 90")
out2<-forecast(lca2,h=20)

par(mfrow=c(2,1))
plot(forecast(lca1,h=20,jumpchoice="fit"),ylim=c(-11,1) )
lines(log(lca1$male[, 1]))

plot(forecast(lca2,h=20,jumpchoice="fit"),ylim=c(-11,1))
lines(lca2$female[, 1])

par(mfrow=c(1,1))
```

# Mechanics of Lee - Carter
I got this example largely from [here](http://rstudio-pubs-static.s3.amazonaws.com/466062_fda07a1b42af4793b1f75e4b87435e3a.html). It uses the `StMoMo` [library ](https://cran.r-project.org/web/packages/StMoMo/vignettes/StMoMoVignette.pdf). You should also totally check out [this video by the author of that package](https://www.youtube.com/watch?v=prk-0G689GU), as to how great it is, also as to why it's called StMoMo.

```{r}
usdat2<-hmd.mx(country = "USA", username = myusername, password = mypassword, label="USA")
usdat2_m <- StMoMoData(usdat2,series = "male")
usdat2_f <- StMoMoData(usdat2,series = "female")


Years <- usdat2$year
Age <- usdat2$age

m <- usdat2$rate$male
m <- log(m)
n.x = nrow(m) # 111
n.y = ncol(m) #78

m_mat <- matrix(m, nrow = n.x, ncol = n.y) # 111 X 85
plot(x=Age, y=m_mat[,1], type = "l", ylim=c(-10, 0), ylab="log rate") #mortality of all ages in year 1940
# Mean log mortality for all age groups under consideration (Age)
a_age <- rowMeans(m_mat)
lines(x = Age, y = a_age, col = 2)#average mortality for all age groups
legend("topleft", legend=c("1940 Mortality - M - Obs","Average Mortality - Male"), col=c(1,2), lty=c(1,1))

```

```{r}
# plotting mortality for Age = 60 as a trial run to see if code is working
# as exxpected
plot(x = Years, y = m_mat[60,], pch = 18, xlab = "Years", ylab = "log m", col = 2,
     main="Age 60 mortality over time") #working!!!
```



```{r}
# LC with StMoMo-----
# Extracting male and female dat from HMD, after creating a StMoMo data object
# "Male" and "Female" data are assigned to different variables for easier
# data wrangling.
library(colorspace)
library(gridExtra)
library(cowplot)
library(RColorBrewer)

#Under a Binomial setting
#Becasue I'm opting to use the data with the Lee-Carter model under a Binomial setting
#the exposures have to be converted to the initial exposures.
LC1 <- lc(link = "logit")
data_m <- central2initial(usdat2_m)
data_f <- central2initial(usdat2_f)
ages_fit <- 20:60


#This can be ussed ot generate a weight matrix over the ages and years in the data.
# clips = 1 assigns 0 weights to the first and last cohorts.
wxt_m <- genWeightMat(ages = ages_fit, years = data_m$years,
                     clip = 1)
wxt_f <- genWeightMat(ages = ages_fit, years = data_f$years,
                      clip = 1)

#For males
LCfit_m <- fit(LC1, data = data_m, ages.fit = ages_fit, wxt = wxt_m)

#For females
LCfit_f <- fit(LC1, data = data_f, ages.fit = ages_fit, wxt = wxt_f)

#plotting parameters
par(mfrow = c(1,3))
plot(x = ages_fit, y = LCfit_m$ax, col = 2, type = "l", ylim=c(-10, -3), main="ax")     #males
lines(x = ages_fit, y = LCfit_f$ax, col = 4)     #females

plot(x = ages_fit, y = LCfit_m$bx, col = 2, type = "l", ylim=c(0, .04), main="bx")
lines(x = ages_fit, y = LCfit_f$bx, col = 4)

plot(x = usdat2_m$years, y = LCfit_m$kt[1,], col = 2, type = "l", ylim=c(-25,40), main="kt")
lines(x = usdat2_m$years, y = LCfit_f$kt[1,], col = 4)
par(mfrow = c(1,1))

```

```{r}
#Goodness-of-fit analysis-----
# For males
res_m <- residuals(LCfit_m)
aic_ <- AIC(LCfit_m)
bic_ <- BIC(LCfit_m)
aic_
bic_
#For females
res_f <- residuals(LCfit_f)

#Plotting colour maps of males and females
p1 <- plot(res_m, type = "colourmap", main = "Residuals of Male data")
p2 <- plot(res_f, type = "colourmap", main = "Residuals of Female data")

```



### Ok, actual forecast
```{r}
#Forecasting----
LCfore_m <- forecast(LCfit_m, h = 50)
LCfore_f <- forecast(LCfit_f, h = 50) 

## Comparison of forcasting done in three instances:
# a.) Forecasting kt with random walk using the forecast funciton.
# b.) Forecast of kt done with SVD and first principles.
# c.) Forecast of kf done with forecast and iarima.
par(mfrow=c(1,2))
plot(x = LCfit_m$years, y = LCfit_m$kt[1,], type = "l",
     ylim=c(-60, 30), xlim=c(min(usdat2_m$years),max(LCfore_m$years)), 
     main="kt for males - forecast",
      ylab = "kt")
lines(x = LCfore_m$years, y = LCfore_m$kt.f$mean, col = 2)
lines(x = LCfore_m$years, y = LCfore_m$kt.f$upper[1,,1], col = 4)
lines(x = LCfore_m$years, y = LCfore_m$kt.f$lower[1,,1], col = 4)

plot(x = LCfit_m$years, y = LCfit_f$kt[1,], xlab = "Years",type = "l",
     ylim=c(-60, 30), xlim=c(min(usdat2_m$years),max(LCfore_m$years)),
     main="kt for females - forecast",
     ylab = "kt")
lines(x = LCfore_m$years, y = LCfore_f$kt.f$mean, col = 2)
lines(x = LCfore_m$years, y = LCfore_f$kt.f$upper[1,,1], col = 4)
lines(x = LCfore_m$years, y = LCfore_f$kt.f$lower[1,,1], col = 4)

```


```{r, fig.height=8, fig.width=12}
LCfit <- fit(lc(link = "logit"), data = usdat2_m, ages.fit = 30:80)

LCfor<-forecast(LCfit, h=50)
plot(LCfor)

```

#More simulations
```{r, fig.height=8, fig.width=12}
LCsim.mrwd <- simulate(LCfit, nsim = 100)
LCsim.iarima <- simulate(LCfit, nsim = 100, kt.method = "iarima")


par(mfrow=c(2, 2))
plot(LCfit$years, LCfit$kt[1, ], xlim = range(LCfit$years, LCsim.mrwd$kt.s$years),
     ylim = range(LCfit$kt, LCsim.mrwd$kt.s$sim), type = "l", 
     xlab = "year", ylab = "kt", 
     main = "Lee-Carter: Simulated paths of the period index kt (mrwd)")
matlines(LCsim.mrwd$kt.s$years, LCsim.mrwd$kt.s$sim[1, , ], type = "l", lty = 1)

plot(LCfit$years, (LCfit$Dxt / LCfit$Ext)["65", ], 
     xlim = range(LCfit$years, LCsim.mrwd$years),
     ylim = range((LCfit$Dxt / LCfit$Ext)["65", ], LCsim.mrwd$rates["65", , ]), 
     type = "l", xlab = "year", ylab = "rate", 
     main = "Lee-Carter: Simulated mortality rates at age 65")
matlines(LCsim.mrwd$years, LCsim.mrwd$rates["65", , ], type = "l", lty = 1)

plot(LCfit$years, LCfit$kt[1, ], xlim = range(LCfit$years, LCsim.iarima$kt.s$years),
     ylim = range(LCfit$kt, LCsim.iarima$kt.s$sim), type = "l", 
     xlab = "year", ylab = "kt", 
     main = "Lee-Carter: Simulated paths of the period index kt (ARIMA(0,1,0))")
matlines(LCsim.iarima$kt.s$years, LCsim.iarima$kt.s$sim[1, , ], type = "l", lty = 1)

plot(LCfit$years, (LCfit$Dxt / LCfit$Ext)["65", ], 
     xlim = range(LCfit$years, LCsim.iarima$years),
     ylim = range((LCfit$Dxt / LCfit$Ext)["65", ], LCsim.iarima$rates["65", , ]), 
     type = "l", xlab = "year", ylab = "rate", 
     main = "Lee-Carter: Simulated mortality rates at age 65 (ARIMA(0,1,0))")
matlines(LCsim.iarima$years, LCsim.iarima$rates["65", , ], type = "l", lty = 1)

```
